For submission to the Journal of Integrated Pest Management

### 2021-02-24 ofv_fife_2021.Rmd
Work done on work PC

### Abstract
Flat mites from the genus *Brevipalpus* Donnadieu (Trombidiformes: Tenuipalpidae) are the only known vectors of dichorhaviruses and transmit *Orchid fleck dichorhavirus* (OFV) in a persistent propagative manner. OFV is the type species for the genus *Dichorhavirus* and infects 50+ spp. of Orchidaceae, Nolinoideae, and *Citrus*.
During June 2020, we observed chlorotic ringspot symptoms on Giant Lilyturf (*Liriope* spp., cv. ‘Gigantea’) in the landscape of Leon County, Florida. We tested plant samples for a wide range of pathogens and ultimately identified the pathogen as *Orchid fleck dichorhavirus*. We took further samples from other *Liriope* spp., *Ophiopogon* spp., and *Aspidistra elatior* Blume (Asparagales: Nolinoideae) in Leon and Alachua county. All symptomatic plants tested positive for OFV. We recovered *Brevipalpus* mites from OFV-infected plants, which belong to the *Brevipalpus californicus* group, but may be a new subspecies associated with Nolinoideae.
Florida has various pest species of *Brevipalpus* and a diverse array of susceptible plant species in the landscape, both native and introduced. We suggest that OFV has a wide distribution in Florida which will continue to spread if unchecked, representing a potential threat for many plants of economic importance for Florida.

@@

On June 2020, chlorotic ringspot symptoms were observed on the liriopogon Giant Lilyturf *Liriope* spp., cv. 'Gigantea' in the landscape of Leon County, Florida (see *`r citef("fig_1")`*). Liropogons belong to the plant family Asparagaceae, subfamily Nolinoideae, which includes various monocotyledonous lilliod plants [@Chase2009]. Liriopogons are native to southeastern Asia, encompassing the closely-related genera *Liriope* and *Ophiopogon*, which are characterized by evergreen, long grass-like leaves, and erect flowers which produce black, purple or blue, berry-like fruits, depending on the cultivar [@Lattier2014; @Fantz2015]. These characteristics have given rise to various common names, including: monkey grass, Aztec grass, lilyturf, mondo grass, etc. (see *`r citet("tab_1")`*). The genus *Liriope* contains eight species, four of which are commonly cultivated, while the genus *Ophiopogon* consists of about 65 species, with only a handful which are commonly cultivated [@Lattier2014; @Nesom2010]. Liripogons are considered the most important ground cover sold by the nursery industry in southeastern US [@Mcharo2003].

Plant samples were initially tested at the Plant Disease Diagnostic Clinic at the North Florida Research and Education Center (NFREC) in Quincy, FL. The plant samples were tested for a wide range of pathogens, including **[Ask Dr. Iriarte]** but the results were inconclusive. The infected materials were subsequently sent to the Florida Department of Agriculture and Consumer Services (FDACS). The pathogen was identified as *Orchid fleck dichorhavirus* via RT-qPCR with a partial sequence match, following the procedures described by @Kubo2009b. **[Which primers were used?]** Further samples were taken from other *Liriope* spp., *Ophiopogon* spp., as well as *Aspidistra elatior* Blume (Asparagales: Nolinoideae) in Leon county. All symptomatic plants were tested positive for OFV. **[Mention the different strains?]**

More cases of OFV-infected Nolinoideae were observed in subsequent visits to the initial site of collection. After a brief search, it appears that the disease is widely distributed in Florida, with symptomatic plants found in Leon, Alachua and Duval counties.

In addition to the discovery of the OFV-infected plants, our lab collected mites from symptomatic plants in Leon county and observed with phase contrast microscopy at the NFREC. We encountered both eriophyoid mites and flat mites which were initially identified as *Brevipalpus californicus* (Banks) *sensu lato*. *Brevipalpus* mites have been previously associated with OFV [@BeltranBeltran2020; @GarciaEscamilla2018; @Dietzgen2018b] and similar diseases [@Kitajima2010] and are known to feed on a large variety of economically-important plants [@Childers2003a; @Akyazi2017]. To verify our identification, we sent the flat mite samples to the FDACS. FDACS used the more advanced techniques of Differential Interference Contrast (DIC) microscopy to verify that our mites belonged to the *B. californicus* group. Unfortunately, even this verification needed further validation, because the *Brevipalpus* mite species complex is known to contain cryptic species [@Childers2011] whose identification can be improved with molecular methods as well as more advanced microscopy techniques, such as low-temperature scanning electron microscopy (LT-SEM) [@Skoracka2015; @Leon2010]. With that in mind, we sent additional samples of the mites to the USDA-ARS in Beltsville, where LT-SEM techniques revealed these mites as a possibly new subspecies of the *B. californicus* group (see *`r citef("fig_2")`*).

The earliest mention of a virus which may have been OFV is @Ko1985, who detected nuclear inclusions caused by an undescribed bacilliform rhabdovirus. @Ko1985 described the spoke-wheel configurations typically associated with OFV (see [@Chang1976]), in *Brassia* orchids, but unfortunately made no mention of mites or further investigations of this virus.

The first certain report of OFV in the US was made by @Bratsch2015, who confirmed the presence of OFV in *Phalaenopsis* hybrids in the US, using TEM of ultrathin sections of plant tissue as well as molecular sequence analysis. In addition, @Bratsch2015 detected *Brevipalpus* mites and their exoskeletons associated with OFV-infected plants. It appears that @Bratsch2015 did not make a conclusive species identification, but did cite @Kondo2003, suggesting that they suspected *B. californicus* as their vector.

OFV has been reported in other Nolinoideae in Australia [@Dietzgen2018b; @Mei2016], including *Liriope spicata* (Thunb.) Lour. [@Mei2016]. We are not aware of any record which reports OFV infection in *Ophiopogon* plants. Unfortunately, liriopogons species are very similar in appearance and growth habit, with few useful characters used for their classification [@Fantz2008a]. Furthermore, the horticultural industry has created a diverse array of cultivars of these plants, which are often mislabeled [@Fantz2008a]. Aside from the horticultural confusion created by humans, natural hybrids between *Ophiopogon* and *Liriope* have created another source of confusion for phylogeny [@Zhou2009]. Together, these factors make it difficult to differentiate and identify species in the landscape. This taxonomic confusion may be accounted for in the future via sequence comparisons of the OFV-infected plants, but these comparisons are beyond the scope of our current report. Nonetheless, we are confident that ours is the first report of OFV infecting *Aspidistra elatior*; although [@Zheng2013] mentions the association of *B. californicus* with *A. elatior* they make no mention of OFV symptoms in this plant.

Detecting OFV in Florida represents a concern for horticulturalist which grow susceptible plants of economic importance such as *Liriope*, *Ophiopogon*, and other susceptible Asparagales spp. which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including the famous Ghost Orchid, *Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe. Furthermore, OFV represents an obstacle to overcome for the burgeoning interest in cultivating Vanilla in southern Florida [@Chambers2019]. Lastly, some OFV isolates are known to be involved with nuclear types of Citrus Leprosis [@Roy2015a], which may be a cause for concern for the citrus industry. In fact, @Kitajima2011a found that the Citrus Leprosis virus (CiLV) which previously affected Florida citrus was a nuclear type of citrus leprosis, which are closely related to OFV strains [@Roy2013a]. *B. californicus* and *B. yothersi* are both known vectors of Dichorhaviruses (OFV) and Cileviruses (Citrus Leprosis) [@BeltranBeltran2020; @Kondo2003; @Knorr1968] and *B. obovatus* is a suspected vector as well [@Childers2003a]. All three mite species/complexes are present in Florida [@Childers2003a; @Akyazi2017] (see *`r citef("fig_3")`*), therefore it is critical to monitor the spread of Orchid Fleck Virus and its mite vector(s) to determine what risk this virus represents for plants in Florida and the surrounding regions.

@@

```{r, fig_1, warning=FALSE}
grid::grid.raster(tiff::readTIFF( "images/fig_1.tif"))
```
`r fig_nums(name = "fig_1", caption = "Variety of symptoms expressed by plants infected with Orchid fleck dichorhavirus: (a) ringspot symptoms on *Liriope gigantea* (b) chlorosis on *Aspidistra elatior*.")`

```{r, fig_2, warning=FALSE}
grid::grid.raster(tiff::readTIFF( "images/b_cali.tif"))
```
`r fig_nums(name = "fig_2", caption = "LT-SEM images of *Brevipalpus californicus* sensu lato used for identification (a) Dorsum  (b) Lateral view (c) Venter (d) Close up of distal end of leg 2, with paired solenidia (e) Enlargement of the microplates of the mite cerotegument (f) Dorsal view of the distal portion of mite abdomen (g) Dorsal view of the mite rostrum (h) Ventral view of mite rostrum LT-SEM images provided by Dr. Gary Bauchan, USDA-ARS 2021")`

```{r, fig_3, warning=FALSE}
grid::grid.raster(tiff::readTIFF( "images/other_mites.tif"))
```
`r fig_nums(name = "fig_3", caption = "Florida is home to other common pest species of *Brevipalpus* which are potential vectors of *Orchid fleck dichorhavirus*: (a) *B. yothersi*, dorsal (b) *B. yothersi*, lateral (c) *B. obovatus*, dorsal")`

@@

### Acknowledgements
We thank Dr. Sam Bolton for confirmation of the mite group.
We are especially indebted to the late Dr. Gary Bauchan for permitting us to use these beautiful LT-SEM images, he will be greatly missed. This research was partly funded by the USDA National Institute of Food and Agriculture, Hatch project FLA-NFC-005607.

>>>>

### Abstract
Flat mites from the genus *Brevipalpus* Donnadieu (Trombidiformes: Tenuipalpidae) are the only known vectors of dichorhaviruses and transmit *Orchid fleck dichorhavirus* (OFV) in a persistent propagative manner. OFV is the type species for the genus *Dichorhavirus* and infects 50+ spp. of Orchidaceae, Nolinoideae, and *Citrus*.
During June 2020, we observed chlorotic ringspot symptoms on Giant Lilyturf (*Liriope* spp., cv. ‘Gigantea’) in the landscape of Leon County, Florida. We tested plant samples for a wide range of pathogens and ultimately identified the pathogen as *Orchid fleck dichorhavirus*. We took further samples from other *Liriope* spp., *Ophiopogon* spp., and *Aspidistra elatior* Blume (Asparagales: Nolinoideae) in Leon and Alachua county. All symptomatic plants tested positive for OFV. We recovered *Brevipalpus* mites from OFV-infected plants, which belong to *Brevipalpus californicus* sensu lato, associated with Nolinoideae.
Florida has various pest species of *Brevipalpus* and a diverse array of susceptible plant species in the landscape, both native and introduced. We suggest that OFV has a wide distribution in Florida which will continue to spread if unchecked, representing a potential threat for many plants of economic importance for Florida.

@@

On June 2020, chlorotic ringspot symptoms were observed on the liriopogon Giant Lilyturf *Liriope* spp., cv. 'Gigantea' in the landscape of Leon County, Florida (see *`r citef("fig_1")`*). Liropogons belong to the plant family Asparagaceae, subfamily Nolinoideae, which includes various monocotyledonous lilliod plants [@Chase2009]. Liriopogons are native to southeastern Asia, encompassing the closely-related genera *Liriope* and *Ophiopogon*, which are characterized by evergreen, long grass-like leaves, and erect flowers which produce black, purple or blue, berry-like fruits, depending on the cultivar [@Lattier2014; @Fantz2015]. These characteristics have given rise to various common names, including: monkey grass, Aztec grass, lilyturf, mondo grass, etc. (see *`r citet("tab_1")`*). The genus *Liriope* contains eight species, four of which are commonly cultivated, while the genus *Ophiopogon* consists of about 65 species, with only a handful are available for horticulture [@Lattier2014; @Nesom2010]. Liripogons are considered the most important ground cover sold by the nursery industry in southeastern US [@Mcharo2003].

Plant leaf samples were initially tested at the Plant Disease Diagnostic Clinic at the North Florida Research and Education Center (NFREC) in Quincy, FL. The plant samples were tested for a wide range of pathogens, including **[Ask Dr. Iriarte]** but the results were inconclusive. The infected materials were subsequently sent to the Florida Department of Agriculture and Consumer Services (FDACS). The presence of *Orchid fleck dichorhavirus* was confirmed via generic one step conventional RT-PCR with R2-Dicho-GF and R2-Dicho-GR primers, amplifying ~800 nt of L-gene (RNA2) amplicon from a leaf sample of infected *Liriope* [@Kubo2009b]. RT-qPCR confirmed OFV and the RT-PCR product was sequenced using Sanger sequencing and High Throughput Sequencing [@Kondo2006]. Sanger sequencing of RT-PCR amplicons shared 98 \% nucleotide identity with OFV strains which infect orchids (GenBank Accession numbers: AB244418 and MK522807) [@Kondo2006; @Kondo2014; @Cook2019]. Further samples were taken from various symptomatic cultivars of *Liriope* spp., *Ophiopogon* spp., as well as *Aspidistra elatior* Blume (Asparagales: Nolinoideae) during subsequent visits to the initial site of collection as well as other locations in Leon county. Samples were tested via RT-PCR at the NFREC (see *`r citet("tab_1")`*) to confirm the presence of OFV.

Further surveys of Nolinoideae in Florida have revealed more sites with symptomatic plants in both Leon and Alachua counties. In addition to the discovery of the OFV-infected plants, our lab at the NFREC collected mites from symptomatic plants in Leon county which were observed with phase contrast microscopy. We encountered both eriophyoid mites and flat mites which were identified as *Brevipalpus californicus* (Banks) sensu lato, an identity which was confirmed by the FDACS via Differential Interference Contrast (DIC) microscopy. *Brevipalpus* mites have been previously associated with OFV [@BeltranBeltran2020; @GarciaEscamilla2018; @Dietzgen2018b] and similar diseases [@Kitajima2010] and are known to feed on a large variety of economically-important plants [@Childers2003a; @Akyazi2017]. Unfortunately, the *Brevipalpus* mite species complex is known to contain cryptic species [@Childers2011] whose identification can be improved with molecular methods as well as more advanced microscopy techniques, such as low-temperature scanning electron microscopy (LT-SEM) [@Skoracka2015; @Leon2010]. With that in mind, we sent additional samples of the mites to the USDA-ARS in Beltsville to observe the mites with LT-SEM techniques, which agreed with both prior identifications of *B. californicus* s.l. (see *`r citef("fig_2")`*).

The earliest mention of a virus which may have been OFV is @Ko1985, who detected nuclear inclusions caused by an undescribed bacilliform rhabdovirus. @Ko1985 described the spoke-wheel configurations typically associated with OFV (see [@Chang1976]), in *Brassia* orchids, but unfortunately made no mention of mites or further investigations of this virus. The first certain report of OFV in the US was made by @Bratsch2015, who confirmed the presence of OFV in *Phalaenopsis* hybrids in the US, using TEM of ultrathin sections of plant tissue as well as molecular sequence analysis. In addition, @Bratsch2015 detected *Brevipalpus* mites and their exoskeletons associated with OFV-infected plants. It appears that @Bratsch2015 did not make a conclusive species identification, but did cite @Kondo2003, suggesting that they suspected *B. californicus* as their vector.

OFV has been reported in other Nolinoideae in Australia [@Dietzgen2018b; @Mei2016], including *Liriope spicata* (Thunb.) Lour. [@Mei2016]. Our plants are thought to belong to different species of *Liriope* which are not *L. spicata* and we are not aware of any record which reports OFV infection in *Ophiopogon* plants. Unfortunately, liriopogons species are very similar in appearance and growth habit, with few useful characters used for their classification [@Fantz2008a]. Furthermore, the horticultural industry has created a diverse array of cultivars of these plants, which are often mislabeled [@Fantz2008a]. Aside from the taxonomic confusion created by humans, natural hybrids between *Ophiopogon* and *Liriope* have created a natural source of error for reconstructing phylogenies [@Zhou2009]. Together, these factors make it difficult to differentiate and identify plants in the landscape by visual inspection alone. These obfuscations of species identity may be accounted for in the future via sequence comparisons of the OFV-infected plants, but these comparisons are beyond the scope of our current report. Nonetheless, we are confident that ours is the first report of OFV infecting *Aspidistra elatior*; although [@Zheng2013] mentions the association of *B. californicus* with *A. elatior* they make no mention of OFV symptoms in this plant.

Detecting OFV in Florida represents a concern for horticulturalists who grow susceptible plants of economic importance, including orchids, *Liriope*, *Ophiopogon*, or other susceptible Asparagales spp. which are commonly used in landscaping. Florida is also home to a plethora of native and naturalized orchid species, many of which are threatened, including the famous Ghost Orchid, *Dendrophylax lindenii* (Lindl.) Benth. ex Rolfe. Furthermore, OFV represents an obstacle to overcome for the burgeoning interest in cultivating *Vanilla* in southern Florida [@Chambers2019]. Lastly, some OFV isolates are known to be involved with nuclear types of Citrus Leprosis [@Roy2015a], which may be a cause for concern for the citrus industry. In fact, @Kitajima2011a found that the Citrus Leprosis virus (CiLV) which previously affected Florida citrus was a nuclear type of citrus leprosis, which are closely related to OFV strains [@Roy2013a]. *B. californicus* and *B. yothersi* are both known vectors of Dichorhaviruses (OFV) and Cileviruses (Citrus Leprosis) [@BeltranBeltran2020; @Kondo2003; @Knorr1968] and *B. obovatus* is a suspected vector as well [@Childers2003a]. All three mite species/complexes are present in Florida [@Childers2003a; @Akyazi2017] (see *`r citef("fig_3")`*), therefore it is critical to monitor the spread of Orchid Fleck Virus and its mite vector(s) to determine what risk this virus represents for plants in Florida and the surrounding regions.

@@

```{r, fig_1, warning=FALSE}
grid::grid.raster(tiff::readTIFF( "images/fig_1.tif"))
```
`r fig_nums(name = "fig_1", caption = "Variety of symptoms expressed by plants infected with Orchid fleck dichorhavirus: (a) ringspot symptoms on *Liriope gigantea* (b) chlorotic flecking on *Aspidistra elatior*.")`

```{r, fig_2, warning=FALSE}
grid::grid.raster(tiff::readTIFF( "images/b_cali.tif"))
```
`r fig_nums(name = "fig_2", caption = "LT-SEM images of *Brevipalpus californicus* sensu lato displaying various characters used for identification [@Baker1987; @Beard2012] (a) Dorsum  (b) Lateral view (c) Venter (d) Close up of distal end of leg 2, with arrows indicating paired solenidia, characteristic of the genus *Brevipalpus* (e) Enlargement of the microplates of the mite cerotegument (f) Dorsal view of the distal portion of mite abdomen (g) Dorsal view of the mite rostrum (h) Ventral view of mite rostrum, observe 3 distal setae. LT-SEM images provided by Dr. Gary Bauchan, USDA-ARS 2021")`

```{r, fig_3, warning=FALSE}
grid::grid.raster(tiff::readTIFF( "images/other_mites.tif"))
```
`r fig_nums(name = "fig_3", caption = "Florida is home to other common pest species of *Brevipalpus* which are potential vectors of *Orchid fleck dichorhavirus*: (a) *B. yothersi*, dorsal (b) *B. yothersi*, lateral (c) *B. obovatus*, dorsal. SEM images provided by Dr. Ron Ochoa, USDA-ARS 2021")`

@@

### Acknowledgements
We would like to give a special thanks to the Tallahassee Museum for their patience, cooperation and support with collecting plant samples. We also want to thank Dr. Sam Bolton for checking the mites we have sent to FDACS for species validation. We are especially indebted to the late Dr. Gary Bauchan for permitting us to use these beautiful LT-SEM images, he will be greatly missed. This research was partly funded by the USDA National Institute of Food and Agriculture, Hatch project FLA-NFC-005607.
